---
title: "Are Tuition Costs Efiecient"
author: "Patrick Chilufya"
date: "May 16, 2018"
output: html_document
---
## Introduction
In this tutorial, we will examine the cost of tuition and average income by state. The goal is to answer the question, "are tuition costs effieecient?". Is investing in college worth it? 

## Part 1: Data scraping and cleaning
We will begin data collection by scraping the College Tuition Compare, the US department of education, wikipedia and the Bureau of Labor Statistics


```{r setup, include=TRUE}
library(rvest) # library need to scrape (or harvest) data from html web page
```

```{r nces_data_scrape }
url <- "https://www.collegetuitioncompare.com/state/"

# .tabletop is a class selector containing the table data we need.
nces_table <- url %>% 
  read_html() %>%
  html_node("table")
nces_table
```
# We now extract the rows of the table that have the data we are interested in
```{r get_rows}
# using a regular expression [scope=\"row\"] to match and extract rows of interest
states_rows <- nces_table %>% html_nodes("a")  %>%
  html_text()
states_rows[60] <- "Total"
states_rows
```

```{r fees}
in_state_tuition_tab <- nces_table %>% html_nodes("tr") %>% html_nodes("td:nth-of-type(2)") %>%
  html_text()

# tuition fee in 2015
off_camp_tab <- nces_table %>% html_nodes("tr") %>% html_nodes("td:nth-of-type(6)") %>%
  html_text()

on_camp_tab <- nces_table %>% html_nodes("tr") %>% html_nodes("td:nth-of-type(5)") %>%
  html_text() %>% as.character()
on_camp_tab
```

```{r tuition_df}
library(utils)
tuition_df <- data.frame(State=states_rows, Tuition_=in_state_tuition_tab, On_Camp_Cost=on_camp_tab, off_camp_cost=off_camp_tab)
#tuition_df <- tuition_df[-c(1), ]
tuition_df$State <- as.character(tuition_df$State)
tuition_df$Tuition <- as.character(tuition_df$Tuition)
tuition_df$Tuition <- gsub('[$,]','',tuition_df$Tuition)
tuition_df$Tuition <- as.numeric(gsub('[$,]','',tuition_df$Tuition))
tuition_df$On_Camp_Cost <- as.numeric(gsub('[$,]','',as.character(tuition_df$On_Camp_Cost)))
tuition_df$off_camp_cost <- as.numeric(gsub('[$,]','',as.character(tuition_df$off_camp_cost)))

tuition_df
```

## Second data scrape

```{r graduates}
library(dplyr)
library(data.table)
url <- "https://www.ed.gov/news/press-releases/new-state-state-college-attainment-numbers-show-progress-toward-2020-goal"

graduate_tab <- url %>% read_html() %>% 
  html_node("table") %>% html_table() %>%
  as.data.frame()

#colnames(graduate_tab) <- as.character(unlist(graduate_tab[1,])) 
#graduate_tab <- graduate_tab[-1, ]
graduate_tab <- graduate_tab %>% select(X1, X6)
graduate_tab <- graduate_tab[-1, ]
setnames(graduate_tab, old=c("X1","X6"), new=c("State", "graduates_range"))

graduate_tab

```

```{r income_by_state}
url <- "https://www.myplan.com/careers/top-ten/highest-paid-states.php"
income_df <- url %>% read_html() %>% html_nodes(".tool_description") %>%
  html_text()
#income_df <- income_df[22] 
#income_df <- income_df %>% html_nodes("tr")
#income_df <- income_df[3] %>% html_nodes("tr")
#income_df <- income_df[1] %>% html_nodes("tr")
income_df <- tail(income_df, -8)
income_df <- as.data.frame(income_df)
state_names <- income_df[seq(1, nrow(income_df), 2), ]

incom_by_state<- income_df[seq(2, nrow(income_df), 2), ]


income_df <- data.frame(State=state_names, income_=incom_by_state)
income_df$State <- as.character(income_df$State)
income_df$income <- as.numeric(gsub('[$,]','',as.character(income_df$income)))

income_df
```

```{r labor_force}
url <- "https://www.bls.gov/news.release/laus.t01.htm"
states_df <- url %>% read_html() %>% html_node("table")  %>% html_nodes("p") %>%
  html_text()

states_df <- head(states_df, 59)

num_jobs_df <- url %>% read_html() %>% html_node("tbody") %>% html_nodes("tr") %>%
  html_nodes("td:nth-of-type(1)") %>%
  html_text()

labor_df <- data.frame(State=states_df, num_jobs=num_jobs_df )
labor_df$State <- as.character(labor_df$State)
labor_df$num_jobs <- as.numeric(gsub('[,]','',as.character(labor_df$num_jobs)))
labor_df
```

```{r final_data_frame}
final_df <-full_join(tuition_df, graduate_tab, by="State") %>% full_join(labor_df, by="State") %>%
  full_join(income_df, by="State")
final_df
```


```{r plot}
library(ggplot2)
plot_df <- select(final_df, State, num_jobs, graduates_range) %>% tidyr::separate(graduates_range, c("Graduate_2018", "Graduate_2020"), "[-]")
plot_df$Graduate_2018 <- as.numeric(gsub('[,]', '', as.character(plot_df$Graduate_2018)))
plot_df$Graduate_2020 <- as.numeric(gsub('[,]', '', as.character(plot_df$Graduate_2020)))
plot_df %>% ggplot(aes(x=Graduate_2018, y=num_jobs)) + geom_smooth() +
  labs(title="Number of Jobs vs Number of Graduates",
         y = "Jobs",
         x = "Graduates")
```



## Part 2: Data wrangling and EDA
We will not standardize tuition cost for both on campus and off campus students for a period of 4 years

```{r standardize_tuition}
library(ggplot2)
standardized_df <- final_df %>%
  mutate(avg_camp_tuiton = mean(Tuition + On_Camp_Cost, na.rm = TRUE)*4, avg_off_camp_tuition = mean(Tuition+off_camp_cost, na.rm = TRUE)*4, four_year_income = income*4) %>%
  mutate(sd_onca_tuition=sd(Tuition+On_Camp_Cost, na.rm = TRUE), sd_offca_tuition = sd(Tuition+off_camp_cost, na.rm = TRUE)) %>%
  mutate(z_oct = (((Tuition+On_Camp_Cost)*4)-avg_camp_tuiton)/sd_onca_tuition) %>%
  mutate(z_ofct = (((Tuition+off_camp_cost)*4)-avg_off_camp_tuition)/sd_offca_tuition)%>%
  select(State, four_year_income, z_oct, z_ofct)
  
standardized_df

```

```{r plot_standardized_data} 
standardized_df %>% ggplot(aes(x=four_year_income, y=z_oct)) + 
  geom_point() +geom_smooth()+
  ylab("Standardized Tuition") +
  xlab("Four Year Income") +
  ggtitle("Standardized Tuition vs Four year Income")
```

## Regression
```{r for_reg}
library(tidyr)
library(ggplot2)
url <- "https://en.wikipedia.org/wiki/List_of_U.S._states_by_income"

new_incomes <- url %>% read_html() %>% html_node(".wikitable") %>% html_table()
new_incomes <- new_incomes %>% gather(year, value=Income, '2015':'2011') 
new_incomes$year <- as.numeric(new_incomes$year)
new_incomes$Income <- as.numeric(gsub('[$,]', '',new_incomes$Income))
new_incomes
```

```{r plot_income_accross_years}
ggplot(data=new_incomes, aes(x=factor(year), y=Income)) + geom_violin()
```

```{r linear_regression_fit}
library(tidyr)
library(gapminder)
auto_fit <- lm(Income~year, data =new_incomes )
auto_fit %>% broom::tidy()
```

```{r inncome_vs_year_plot}
new_fit_df <- lm(Income~year, data =new_incomes ) %>%
broom::augment() %>%
inner_join(new_incomes, by=c('year', 'Income'))

new_fit_df %>%
ggplot(aes(x=State, y=.resid)) +
geom_point() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
labs(title="Model redidual vs State ",x="State", y="model redidual")
```

```{r linear_model_fit}
new_fit_df %>% ggplot(aes(x = year, y = Income, color=State)) +
geom_point() +
geom_smooth(method=lm)+
labs(title="Income vs Year ",x="Year", y="Income")
```

```{r linear_regression_model_with_state}
interaction_fit <- lm(Income~year*State, data =new_incomes )
interaction_fit %>% broom::tidy()
```
## Classification
For classification, we will use the new_incomes data frame created in the regression
```{r classification}
library(tidyr)
library(dplyr)

stdzd_df <- new_incomes %>% group_by(State) %>%
  mutate(mean_inc = mean(Income)) %>%
mutate(sd_inc = sd(Income)) %>%
mutate(z_inc = (Income - mean_inc) / sd_inc) %>%
ungroup()

wide_df <- stdzd_df %>% select(State, year, z_inc) %>% spread(year,z_inc)
predictor_df <- new_incomes %>% spread(year, Income) 
predictor_df <- predictor_df %>% mutate(diff=predictor_df$`2015`- predictor_df$`2014`) %>%
  mutate(Direction=ifelse(diff>0, "up", "down"))
predictor_df

```


```{r quarterly_differences}

matrix_1 <- wide_df %>%
  select(-State) %>%
  as.matrix() %>%
  .[,-1]


matrix_2 <- wide_df %>%
select(-State) %>%
as.matrix() %>%
.[,-ncol(.)]


diff_df <- (matrix_1 - matrix_2) %>%
magrittr::set_colnames(NULL) %>%
as_data_frame() %>%
mutate(State = wide_df$State)


classifier_df <- diff_df %>%
inner_join(predictor_df %>% select(State, Direction), by="State") %>%
mutate(Direction=factor(Direction, levels=c("down", "up")))


classifier_df
```

```{r random_forest}
library(randomForest)
set.seed(1234)
test_random_forest_df <- classifier_df %>%
  group_by(Direction) %>%
  sample_frac(.2) %>%
  ungroup()

train_random_forest_df <- classifier_df %>%
anti_join(test_random_forest_df, by="State")

rf <- randomForest(Direction~., data=train_random_forest_df %>% select(-State))
rf
```

```{r predictions}
test_predictions <- predict(rf, newdata=test_random_forest_df %>% select(-State))
table(pred=test_predictions, observed=test_random_forest_df$Direction)

```

```{r AUROC_curve}
library(ROCR)
library(ISLR)
library(cvTools)

standardzd_pred <- predict(rf, type="prob", newdata = classifier_df %>%select(-State))
stadardzd_prediciton <- prediction(standardzd_pred[,"up"], classifier_df$Direction) 
auc <- unlist(performance(stadardzd_prediciton,"auc")@y.values)
plot(performance(stadardzd_prediciton, "tpr", "fpr"), main=paste("LDA AUROC=", round(auc, 2)), lwd=1.4, cex.lab=1.7, cex.main=1.5)
```


```{r }